The Ranked Nearest Neighbour Interchange space

### 2019-06-26 paper.tex
rm weak split thm, instead: cluster thm

\todo{define splits and delete definition from split thm chapter!}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{RNNI}
	\caption{Four trees connected by edges in the $\rnni$ graph. A rank move of $mrca(\{1,2\})$ and $mrca(\{3,4\})$ and all $\nni$ moves on the edge highlighted in red are illustrated}
	\label{fig:NNI_vs_RNNI}
\end{figure}

@@

Instead of using the notion of clusters as we do within this paper, the Split Theorem considers partitions of the taxon set $X$, so-called splits, received by deleting an edge of a tree and taking the taxon sets of the two resulting trees as partition of $X$.
However, there is a simple counterexample to this conjecture:

@@

Since the counterexample in Figure~\ref{splitthm_counterexample} clearly shows that the version of the Split Theorem stated in~\cite{Gavryushkin2017} does not hold, we will now claim an alternative to this conjecture.
The alternative formulation is based on the observation that there is a shortest path preserving all splits between the trees of the counterexample of the Split Theorem, which is presented at the bottom of Figure~\ref{splitthm_counterexample}.

\begin{conjecture}[Weak Split Theorem]
	For $\rnni$ graph the following statement holds:
	if a partition of leaves given by an edge is present in two trees $T$ and $R$ then there exists a shortest path between $T$ and $R$ where this partition is present in every tree on this path.
	\label{split_theorem_weak}
\end{conjecture}

\todo{Do we want to keep the weak version of the split thm? It is stronger than the Cluster Thm!}

% The weak Split Theorem (in the version as in Lena's thesis) is computationally proven to hold for trees with up to $6$ taxa ($\sim 38$ min running time on laptop).

Another alternative formulation of the Split Theorem is the \emph{Cluster Theorem}, which considers clusters instead of splits.
Considering clusters rather than splits is more reasonable since two rooted trees have the same set of clusters if, and only if, they are isomorphic~\cite{steel2016}.
The trees $T, R$ of counterexample~\ref{splitthm_counterexample} share the same set of induced splits, but have distinct sets of clusters.

@@

>>>>

\todo{define splits and delete definition from split thm chapter!}
\todo{change definition of clusters: induced by vertices, not edges!}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{RNNI}
	\caption{Four trees connected by edges in the $\rnni$ graph. A rank move of $mrca(\{1,2\})$ and $mrca(\{3,4\})$ and all $\nni$ moves on the edge highlighted in red are illustrated}
	\label{fig:RNNI}
\end{figure}

@@

Instead of using the notion of clusters as we do within this paper, the Split Theorem considers partitions of the taxon set $X$, so-called splits, received by deleting an edge of a tree and taking the taxon sets of the two resulting trees as partition of $X$.
However, we found a simple counterexample to this conjecture:

@@

Since the counterexample in Figure~\ref{splitthm_counterexample} clearly shows that the version of the Split Theorem stated in~\cite{Gavryushkin2017} does not hold, we will now claim an alternative to this conjecture, the \emph{Cluster Theorem}.
Within the Cluster Theorem we are considering clusters instead of splits.
This is motivated by the fact that rooted phylogenetic trees can uniquely be represented by sets of clusters~\ref{steel2016}, but not by sets of splits as they cannot define the position of the root.
Moreover, a ranked tree can be uniquely determined by ordering the set of clusters increasingly according to the ranks of the internal nodes inducing those clusters.

% \begin{conjecture}[Weak Split Theorem]
% 	For $\rnni$ graph the following statement holds:
% 	if a partition of leaves given by an edge is present in two trees $T$ and $R$ then there exists a shortest path between $T$ and $R$ where this partition is present in every tree on this path.
% 	\label{split_theorem_weak}
% \end{conjecture}
%
% \todo{Do we want to keep the weak version of the split thm? It is stronger than the Cluster Thm!}
%
% The weak Split Theorem (in the version as in Lena's thesis) is computationally proven to hold for trees with up to $6$ taxa ($\sim 38$ min running time on laptop).

@@

Note that the trees $T, R$ of counterexample~\ref{splitthm_counterexample} share the same set of induced splits, but have distinct sets of clusters.
Furthermore, the example used in $\rnni$ to show that there are trees that share splits that are not present on any shortest path between these trees can be used to prove that the Cluster Theorem does not hold in $\nni$.
